---
title: "Model"
output:
  html_document:
    df_print: paged
---


```{r, eval=TRUE, message=FALSE, warning=FALSE}
# Code imports
library(dplyr)

training_data = readr::read_csv(
                        here::here("data/pml-training.csv"),
                        col_types = readr::cols(
                          .default = readr::col_double(),
                          user_name = readr::col_factor(),
                          classe = readr::col_factor(),
                          raw_timestamp_part_1 = readr::col_integer(),
                          raw_timestamp_part_2 = readr::col_integer(),
                          cvtd_timestamp = readr::col_datetime(format = "%d/%m/%Y %H:%M"),
                          new_window = readr::col_factor()))


training_data = training_data[c(selected_vars, "classe")]
```

## Model 0
The first model will be created using 70% of the training data. It will be used to calculate the variables importances using a Random Forest
```{r}
set.seed(12345)

train_idx  = caret::createDataPartition(training_data$classe, p = 0.7, list=FALSE)

train_control = caret::trainControl(method="cv", number = 10)

model0 = caret::train(classe ~ . , 
                      data = training_data[train_idx, ], 
                      method="rf", 
                      importance = TRUE,
                      trControl = train_control,
                      preProcess= c("center","scale"))

model0
caret::varImp(model0)
```

Looking on their importances, it seems that those variables contributes the most:
```{r}

selected_vars <- c("roll_belt", "pitch_belt", "yaw_belt", "magnet_dumbbell_y", "roll_arm", "magnet_dumbbell_z")

model1 = caret::train(classe ~ ., 
                      data = training_data[train_idx, c(selected_vars, "classe")],
                      method="rf",
                      trControl = train_control,
                      preProcess= c("center","scale"))
model1
```

A taxa de erro está boa, embora um pouco otimista. Vamos ver
The error rate seems good enough with model1, altough they may be a bit optmistic, since the variable selection was done on the same 70% of the data where the cross validation was used.



## Evaluation on the 30%

```{r}
predictions_model1 <- predict(model1, 
                              training_data[-train_idx, c(selected_vars, "classe")])
caret::confusionMatrix(training_data[-train_idx,"classe"]$classe, predictions_model1)
```


# Test Data
```{r}
testing_data = readr::read_csv(
                        here::here("data/pml-testing.csv"),
                        col_types = readr::cols(
                          .default = readr::col_double(),
                          user_name = readr::col_factor(),
                          classe = readr::col_factor(),
                          raw_timestamp_part_1 = readr::col_integer(),
                          raw_timestamp_part_2 = readr::col_integer(),
                          cvtd_timestamp = readr::col_datetime(format = "%d/%m/%Y %H:%M"),
                          new_window = readr::col_factor()))

predictions_test <- predict(model1, testing_data[ , c(selected_vars, "classe")])
readr::write_csv(predictions_test, here::here("outputs/test_results.csv"))
```


